Automatically Splitting a Staged Lambda Calculus

### 2015-07-10 paper/relatedwork.tex
Minor editing on some of the work I did last night.

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose; however, the key difference between our work and this prior
work is that we instead focus on the problem of splitting.
%we developed an algorithm
%for splitting terms in the proposed staged language.
%the staging or the splitting algorithm, which given a two-staged
%program splits it into two stages that can be executed sequentially,
%passing information from the earlier to the latter stage by
%construction.
%
%Unlike partial evaluation, splitting requires no knowledge of program
%argument values, only their staging.
%Using just the staging
%information, the splitting algorithm can generate programs that can
%perform non-trivial transformations on the code automatically.  As our
%examples illustrate such staging transformations can improve the
%efficiency of the source program, sometimes in asymptotically
%significant ways.

Another class of precomputation techniques is incremental computation,
where a program can efficiently respond to small changes in its inputs
by only recomputing parts of the computation affected by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Thus, incremental computation can be thought of as a staged model,
where changes in inputs come over a period of time in stages.
%
However, unlike splitting, incremental computation does not require
fixing any of its inputs and, in the general case, allows for all
program inputs to change. Thus, the benefits of incremental
computation depends on what changes to inputs are made. For example,
while it is possible to apply incremental computation to our
quickselect example (\ref{sec:overview}), techniques would unfold the
quickselect function based on the demanded ranks, potentially
incurring linear time cost at each step of the algorithm (as opposed
to logarithmic result produced by splitting). Unlike the results of
splitting, such an implementation would also maintain sophisticated
data structures dynamically at run time to track what computations
must be performed.

@@

Splitting algorithms have also been a topic of interest in programming
systems for computer graphics, where to achieve high performance,
programs are typically manually separated into components by frequency
of execution. (Indeed, separation into a pre-determined set of
per-frequency modules: e.g., per frame, per pixel, per triangle, is
\emph{required} by the execution model of modern graphics
architectures.)  The software engineering challenges of modifying
graphics programs written in a split form have spurred interest in
deferring the task of splitting to a compiler\,\cite{Foley:2011}.
While early work sought to determine the staging of programs
automatically via binding time analysis\,\cite{knoblock96}, more
recent designs, like \lang, require explicit staging by the
programmer\,\cite{Proudfoot:2001,Foley:2011,He:2014}.  However,
splitting algorithms developed in all of these efforts have been
limited to simple, imperative graphics shading languages that lack
simple features such as functions.

>>>>

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose; however, the key difference between our work and this prior
work is that we instead focus on the problem of splitting.

Another class of precomputation techniques is incremental computation,
where a program can efficiently respond to small changes in its inputs
by only recomputing parts of the computation affected by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Thus, incremental computation can be thought of as a staged model,
where changes in inputs come over a period of time in stages.
However, unlike splitting, incremental computation does not require
fixing any of its inputs and, in the general case, allows for all
program inputs to change. Thus, the benefits of incremental
computation depend on what changes to inputs are made. For example,
while it is possible to apply incremental computation to the
quickselect example in \ref{sec:overview}, techniques would unfold the
quickselect function based on the demanded ranks, potentially
incurring linear time cost at each step of the algorithm (as opposed
to the logarithmic result produced by splitting).  Unlike splitting,
incremental computation techniques must also maintain sophisticated
data structures dynamically at run time to track what computations
must be performed.

@@

Splitting algorithms have also been a topic of interest in programming
systems for computer graphics, where to achieve high performance,
programs are typically manually separated into components by frequency
of execution. (Indeed, separation into a pre-determined set of
per-frequency modules: e.g., per frame, per pixel, per triangle, is
\emph{required} by the execution model of modern graphics
architectures.)  The software engineering challenges of modifying
graphics programs written in a split form have spurred interest in
deferring the task of splitting to a compiler\,\cite{Foley:2011}.
While early work sought to determine the staging of programs
automatically via binding time analysis\,\cite{knoblock96}, more
recent designs, like \lang, require explicit staging by the
programmer\,\cite{Proudfoot:2001,Foley:2011,He:2014}.  However,
splitting algorithms developed in all of these efforts have been
limited to simple, imperative graphics shading languages that lack
even simple features such as functions.
